---
title: "Supplementary text for GBE ms GBE-210319 " 
subtitle: "Hunting for beneficial mutations"
author: "J Chen, T. Bataillon, S Glemin, M. Lascoux"
date: "Last update: `r Sys.Date()`"
output:
  workflowr::wflow_html:
    toc: yes
  # pdf_document:
  #   toc: yes
  #   toc_depth: '2'
  # html_document:
  #   theme: united
  #   code_folding: show
  #   toc: yes
  #   toc_depth: 2
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(ggthemes)
library(dplyr)
library(magrittr)
library(knitr)
library(lme4)
library(merTools)
library(lmerTest)
```

# brief overview of analysis and updates :

* We read in the data made available in the lastest version of the poyDFE outputs summary by Jun Chen.

* We define a covariate $\delta$ as the change in discretized SIFT scores

* Rationale for conditioning DFE on $\delta$ is to illustrate that change in SIFT scores are a powerful way to capture the expected effect of mutations and the fact that DFEs are quite different. There is a sharp divide between $\delta \leq 0$ and $\delta >0$ 

* we explore the amount of variation explained by $\delta$ for 
   + pin/piS
   + other DFE property: probability of sizable amount of beneficial mutations ( eg $Prob[ p_b >0.1 ]$)  


# Reading the data 

```{r data}
dfe_sift <- read.table("Data/summary_polyDFE_siftCate_new.txt",header=T)
dim(dfe_sift)
names(dfe_sift)
dfe_sift$delta <- dfe_sift$to - dfe_sift$from
head( dfe_sift) %>% kable(digits = 2)
```

# Filtering data : 

Removing obs with too high $\epsilon_{anc}$

I n what follows I use the subset obtained with the condition  `filter(cat04==0)` to generate `df_tofit` used throughout for fitting .
NB Filtering on `eps` means 15% data obs removed but I think it is needed in order to avoid weird outliers 

```{r}
dim(dfe_sift)
species_low_nchr <- c("Qmango", "Shua", "Bnana")
dfe_sift <- dfe_sift %>%  
  filter(!(species %in% species_low_nchr)) %>%
  filter(eps <0.1)  # 
dim(dfe_sift)

dim(dfe_sift)
(507-387)/507 #data loss

dfe_sift %>% filter(cat04==0) -> df_tofit
dim(df_tofit)

length(unique(df_tofit$species))

```

# Dependent variable : $\frac{P_0}{P_4}$

We use the corrected $\frac{P_0}{P_4}$ analogous to $\frac{\pi_N}{\pi_S}$

We correct by adding +1 to every count: $\pi_N =\frac{n_N + 1}{L_N + 1}$ and $\pi_S =\frac{n_S + 1}{L_S + 1}$ we can then directly use a log scale

```{r}
df_tofit$PiNPiScor <- (df_tofit$nsyn_counts+1)*(df_tofit$Lsyn+1)/((df_tofit$Lnsyn+1)*(df_tofit$syn_counts+1)) 

df_tofit %>%
  filter(cat04==0) %>%
  ggplot(aes(x=delta, y=PiNPiScor, weight=(1+nsyn_counts)) )+ 
  # geom_point(aes(size = nsyn_counts)) + 
  geom_jitter(width = 0.2, aes(size = nsyn_counts)) + 
  scale_y_log10() +
  # facet_wrap(~ group) + 
  geom_smooth(method = "loess", formula ="y~ x", se =T, aes(color=NULL))+
  geom_smooth(method = "lm", formula ="y~ x", se =T, aes(color=NULL), color="orange")+
  theme_minimal(base_size = 20)+
  xlab(expression(paste("Change in SIFT score, " , delta)))+
  ylab("corrected P0/P4")+
  scale_color_colorblind()+
  theme(legend.position = "none")+
  NULL
# ggsave("manuscript_GBE/manuscript_GBE_Jun2/figures_update_Feb2020/corrected_p0p4_delta.pdf", device = "pdf")
```


## A (linear) fit for each species separately
```{r}
df_tofit %>%
  filter(cat04==0) %>%
  ggplot(aes(x=delta, y=PiNPiScor, color=species, weight=sqrt((1+nsyn_counts))) )+ 
  # geom_point(aes(size = nsyn_counts)) + 
  geom_jitter(width = 0.2, aes(size = nsyn_counts)) + 
  scale_y_log10() +
  # facet_wrap(~ group) + 
  geom_smooth(method = "lm", formula ="y~ x", se =F)+
  theme_minimal(base_size = 20)+
  xlab(expression(paste("Change in SIFT score, " , delta)))+
  ylab("corrected P0/P4")+
  scale_color_viridis_d(direction = -1)+
  theme(legend.position = "none")+
  NULL
# ggsave("manuscript_GBE/manuscript_GBE_Jun2/figures_update_Feb2020/corrected_p0p4_delta_manyslopes.pdf",device = "pdf")
```

## Linear (mixed) models

```{r}

dim(df_tofit)

df_tofit$weights <- sqrt(df_tofit$nsyn_counts)

qplot(log10(df_tofit$PiNPiScor), bins = 25)

# ggplot(df_tofit, 
#        aes(x=delta, y = log10(PiNPiScor), 
#            size = )) + 
#   geom_jitter(width = 0.1)+ 
#   geom_smooth(method="lm", se =T)+ NULL
# 

naive0 <- lm(log(PiNPiScor) ~ 1 , 
            data = df_tofit)
naive <- lm(log(PiNPiScor) ~ delta , 
            data = df_tofit) 
summary(naive)
# plot(naive)

AIC(naive)
mixed_1 <-lmer(formula = log(PiNPiScor) ~ delta +(1|species) , 
     data = df_tofit, REML =F )

merTools::modelInfo(mixed_1)
lmerTest::ranova(mixed_1) 

mixed_2 <-lmer(formula = log(PiNPiScor) ~  delta +(0 + delta |species), 
     data = df_tofit,  REML =F )

AIC(mixed_2)
merTools::modelInfo(mixed_2) 

plotREsim(REsim(mixed_1)) # estimated random effects (intercepts) for each sp

summary(mixed_2)
plotREsim(REsim(mixed_2)) # estimated random effects (slopes) for each sp relative to the common slope
```

## Summary of fits
```{r}
vec_aic = c(AIC(naive0), AIC(naive), AIC(mixed_1), AIC(mixed_2))
models_names = c("lm no slope", "lm slope", "random intercept", "random slope") 
df_table <- tibble(models_names, vec_aic)
knitr::kable(df_table, digits = 2, col.names = c("Models for P0/P4 ratio", "AIC"))
```

```{r}
1 - (deviance(mixed_2) /(1011.67-1))
```

## Tech check: influence of eps
```{r}
naive0 <- lm(log(PiNPiScor) ~ eps , 
            data = df_tofit)
summary(naive0)
naive <- lm(log(PiNPiScor) ~ eps + delta , 
            data = df_tofit) 
summary(naive)

anova(naive, test = "F")

car::vif(naive)
# plot(naive)

AIC(naive)
mixed_1 <-lmer(formula = log(PiNPiScor) ~ eps + delta +(1|species) , 
     data = df_tofit, REML =F )

merTools::modelInfo(mixed_1)
lmerTest::ranova(mixed_1) 

mixed_2 <-lmer(formula = log(PiNPiScor) ~  eps + delta +(0 + delta |species), 
     data = df_tofit,  REML =F )

AIC(mixed_2)
merTools::modelInfo(mixed_2) 

plotREsim(REsim(mixed_1)) # estimated random effects (intercepts) for each sp

summary(mixed_2)
plotREsim(REsim(mixed_2)) # estimated random effects (slopes) for each sp relative to the common slope

vec_aic = c(AIC(naive0), AIC(naive), AIC(mixed_1), AIC(mixed_2))
models_names = c("lm no slope", "lm slope", "random intercept", "random slope") 
df_table <- tibble(models_names, vec_aic)
knitr::kable(df_table, digits = 2, col.names = c("Models for P0/P4 ratio", "AIC"))

```

# Dependent variable : $p_b$
Here if we want to analyse variation in $p_b$, we probably need to discretize the data and go a logistic regression on the two categories 

## Overview 
```{r}
qplot(df_tofit$pb, bins = 30, fill = (df_tofit$pb>0.1)) + 
  theme_minimal(base_size = 15) 
df_tofit <- mutate(df_tofit, benef_mut = ifelse(pb<0.1,  0, 1)) 
table(df_tofit$benef_mut) %>% kable(col.names = c("benef_mut", "n"))
```


```{r}
  df_tofit %>%
  ggplot(aes(x=delta, y=benef_mut, color=species, weight=sqrt((1+nsyn_counts))) )+ 
  # geom_point(aes(size = nsyn_counts)) + 
  geom_jitter( width = 0.15, height = 0.05, aes(size = nsyn_counts)) + 
  geom_smooth(method = "glm",
              method.args=list(family="binomial"), se=T, aes(color="black"), size = 2) + 
  theme_minimal(base_size = 20)+
  xlab(expression(paste("Change in SIFT score, " , delta)))+
  ylab(expression(paste("Probability of " , p[b], " > 0.1")))+
  scale_color_viridis_d()+
  theme(legend.position = "none")+
  NULL
```


## A loess fit for each species separately
## Linear (mixed) models

```{r}
# df_tofit<- df_tofit[-c(247),] # 1 mega outlier
dim(df_tofit)

df_tofit$weights <- sqrt(df_tofit$nsyn_counts)


naive0 <- glm(benef_mut  ~ 1 , family = "binomial",
            data = df_tofit) 
summary(naive0)
naive0$deviance
naive <- glm(benef_mut  ~ delta , family = "binomial",
            data = df_tofit) 
summary(naive)
anova(naive0, naive, test="Chisq")
plot(naive)

1 - (naive$deviance/naive0$deviance) # pseudo R^2 

AIC(naive)
mixed_1 <-glmer(benef_mut ~ delta +(1|species) , family = "binomial",
     data = df_tofit )

AIC(mixed_1)
summary(mixed_1)
merTools::modelInfo(mixed_1)

mixed_2 <-glmer(formula = benef_mut ~  delta +(0 + delta |species), family = "binomial",
     data = df_tofit)

AIC(mixed_2)
summary(mixed_2)
merTools::modelInfo(mixed_2) 

plotREsim(REsim(mixed_1)) # estimated random effects (intercepts) for each sp
plotREsim(REsim(mixed_2)) # estimated random effects (slopes) for each sp relative to the common slope


```

## Summary of fits
```{r}
vec_aic = c(AIC(naive0), AIC(naive), AIC(mixed_1), AIC(mixed_2))
models_names = c("glm no slope", "glm with slope", "glm with random intercept", "glm with random slope") 
df_table <- tibble(models_names, vec_aic)
knitr::kable(df_table, digits = 2, col.names = c("Models for proportion beneficial ", "AIC"))
```

Some pseudo R^2 to compare with the linear models on the piNpiS ratio 
Using the deviance of `mixed_2` and comparing it to the most naive model (glm no slope) aka `naive0` yields a pseudo R^2: 

```{r}
 1 -(deviance(mixed_2)/naive0$deviance)
 
```

# tech check with eps included in the model 
```{r}
naive0 <- glm(benef_mut  ~ eps , family = "binomial",
            data = df_tofit) 
summary(naive0)
naive0$deviance
naive <- glm(benef_mut  ~ eps + delta , family = "binomial",
            data = df_tofit) 
summary(naive)
anova(naive0, naive, test="Chisq")

1 - (naive$deviance/naive0$deviance) # pseudo R^2 

AIC(naive)
mixed_1 <-glmer(benef_mut ~ eps + delta +(1|species) , family = "binomial",
     data = df_tofit )

AIC(mixed_1)
summary(mixed_1)
merTools::modelInfo(mixed_1)

mixed_2 <-glmer(formula = benef_mut ~ eps +  delta +(0 + delta |species), family = "binomial",
     data = df_tofit)

AIC(mixed_2)
summary(mixed_2)
merTools::modelInfo(mixed_2) 

plotREsim(REsim(mixed_1)) # estimated random effects (intercepts) for each sp
plotREsim(REsim(mixed_2)) # estimated random effects (slopes) for each sp relative to the common slope

car::vif(naive)
 df_tofit %>%
  ggplot(aes(x=eps, y=benef_mut, color=species))+ 
  # geom_point(aes(size = nsyn_counts)) + 
  geom_jitter( width = 0.0025, height = 0.05, aes(size = nsyn_counts)) + 
  geom_smooth(method = "glm",
              method.args=list(family="binomial"), se=T, aes(color=NULL), color="black") + 
  theme_minimal(base_size = 15)+
  xlab("eps")+
  ylab("Probability of p_b >0.1 ")+
  scale_color_viridis_d()+
  theme(legend.position = "none")+
  NULL
 
car::vif(naive)
 
 df_tofit %>%
  ggplot(aes(x=delta, y=benef_mut, color=species))+ 
  # geom_point(aes(size = nsyn_counts)) + 
  geom_jitter( width = 0.0025, height = 0.05, aes(size = nsyn_counts)) + 
  geom_smooth(method = "glm",
              method.args=list(family="binomial"), se=T, aes(color=NULL), color="black") + 
  theme_minimal(base_size = 15)+
  xlab(expression(paste("Change in SIFT score, " , delta)))+
  ylab("Probability of p_b >0.1 ")+
  scale_color_viridis_d()+
  theme(legend.position = "none")+
  NULL
 
 # updated pseudo R2
  1 -(deviance(mixed_2)/naive0$deviance)
```




